---
title: "Homework about Indexing and Subsets"
subtitle: "Always ask Will if you need help"
output:
  pdf_document: default
  html_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}

```

```{r}
#rm(list = ls())
```

[1] “apples” “wheat” “soybeans” “potatoes” “grapes” “cherries” “almonds” “hay”
[9] “hops
#Homework Begin
Create a crop vector
```{r}
#Make the vector

crops <- c("apples","wheat","soybeans","potatoes","grapes","cherries","almonds","hay","hops")
replacementcrops <- crops

replacementcrops[3] <- "Oranges"

WAcropswvector <- crops[c(1,5,6,2,9,8)]

WAcropswlogic <- c(TRUE,TRUE,FALSE,TRUE,TRUE,TRUE,FALSE,TRUE,TRUE)

Samevectorasabove <- crops[WAcropswlogic]


order(WAcropswvector)
order(sort(WAcropswvector))

print(sort(WAcropswvector))

```

These are WA crops fyi: apples” “grapes” “cherries” “potatoes” “hops” “wheat” “hay”


#Homework Q 2

```{r}
matdata <- c(1:20)
q2mat <- matrix(data = matdata, ncol = 5, nrow = 4, byrow = TRUE)

```


```{r}
just45 <- c(q2mat[1,4],q2mat[1,5])
just45
```
```{r}
norow3 <- q2mat[-3,]
norow3
```


Let's Talk About Lists



Matrix can only hold 1 atomic data type so 
we need another data structure - Lists can hold different types and objects

We use the List Function

```{r}
vect1 <- c(1:3)
vect2 <- c("Central","Washington","University")
listexample <- list(vect1,vect2)
```



We can even put other lists or a matrix... other multidimensional objects

```{r}
Matex <- matrix(data = c(1:5),nrow = 3)

listexample$object3 <- Matex

listexample
```



```{r}
names(listexample) <- c("text","numbers","multi")
```


```{r}
listexample[1] <- c("Soils 502")

listexample
```

```{r}
listexample[1:2]
```
Double Index

```{r}
listexample[[2]][2]
```



To remove we do index assignment and give it Null

```{r}

listexample[3] <- NULL

listexample$object3 <- Matex

```

```{r}
secondlist <- list(c("Go"),c("Wildcats"))

listexample[[6]] <- secondlist
```


Lists are a good introduction to working with Data frames :)



#CBIND
&
#RBIND


```{r}
### Homework data
##[1] 3 5 6 12
##
## [1] TRUE FALSE TRUE TRUE
##
## [1] "A" "B" "C" "D" "E"
##
## [,1] [,2] [,3] [,4] [,5]
## [1,] 2 14 26 5 17
## [2,] 5 17 29 8 20
## [3,] 8 20 32 11 23
## [4,] 11 23 2 14 26
##
## [,1] [,2] [,3] [,4]
## [1,] 1 -1 1 0
## [2,] 0 0 -1 1
```






```{r}
Q4vect1 <- c(3, 5, 6, 12)

Q4vect2 <- c(TRUE,FALSE,TRUE,FALSE)

Q4vect3 <- c("Mary","Had A","Little","Lamb")

matrow1 <- c(2, 14, 26, 5, 17)
matrow2 <- c(5, 17, 29, 8, 20)
matrwow3 <- c(8,20,32,11,23)
matrow4 <- c(11,23,2,14,26)



Q4mat1 <- matrix(data = rbind(matrow1,matrow2,matrwow3,matrow4),nrow = 4 )

```


cbind time

```{r}
v2 <- cbind(Q4vect1,Q4vect2)

v2
```

```{r}
v3 <- cbind(Q4vect3,Q4vect2)
v3
```
```{r}
result <- cbind(Q4mat1,Q4vect1)
result
result <- cbind(Q4vect1,Q4mat1)
result
```
Rbind

```{r}
#producing the data object from Q4 part c
result <- rbind(Q4mat1,Q4vect1)
result
result <- rbind(Q4vect1,Q4mat1)
result
```


```{r}
rbindexample <- rbind(v2,v3)
rbindexample

print("second example")
rbindexample <- rbind(rbindexample,rbindexample)
rbindexample
```


















































