---
title: "Introduction to dataframes, subsetting and addressing"
subtitle: "Data in R, Fall 2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Factors

**Factors** and **ordered factors** are similar to vectors, but contain categorical data. R handles factors differently than vectors, so a factor is more than vector with a different type of data.

## grades as factors

The **as.factor** function can be used to create a factor from a vector. We can demonstrate with a vector of student grades.
```{r factor chr}
# convert character vector to factor
grades <- c("A", "B", "C", "F", "B", "A", "C", "C", "C", "B", "B", "B", "B", "A", "C", "B", "A", "F")
grades

cat("\nnow create a factor:\n")
grades_fact <- factor(grades)
grades_fact

cat("\nstr\n") # blank line
str(grades_fact)

cat("\nattributes\n") # blank line
attributes(grades_fact)
```

The **factor** is stored in R as 1-dimensional string of levels (1,2,3,4) and those levels have assigned labels of "A", "B", "C" and "D".

A variable should be designated a factor if it is categorical and the number of categories is not too large (far fewer levels than observations.) If there are a large number of categories that are to be used descriptively, then a character vector is more appropriate.

## ordered factor

For ordinal data, we want to use an **ordered factor**.
```{r ordered factor chr}
grades_ord <- factor(grades, ordered = T)
grades_ord
cat("\n")
str(grades_ord)
cat("\nattributes\n") # blank line
attributes(grades_ord)
```

THe order by default is alphabetical, but we probably want A > B > C > F. Also notice that there is no "D" grade. The factor function just works with the levels given. To fix these problems we can set the levels.
```{r ordered factor chr 2}
# set ordered=T to create an ordered factor
grades_ord2 <- factor(grades, ordered = T, levels = c("F", "D", "C", "B", "A"))
grades_ord2
cat("\n")
str(grades_ord2)
cat("\nattributes\n") # blank line
attributes(grades_ord2)
```

Notice that we now have a "D" level even though there are no D grades in the ordered factor.

## Numerical data to factors

R can also convert numerical or integer data to factors. This data is coerced into character data, then the factor function operates as illustrated previously.
```{r factor numerical}
samp <- sample(x = c(1, 3, 4, 7, 9), size = 90, replace = T)
str(samp)

# convert numeric data to a factor
samp_fact <- factor(samp)
cat("\nPrint factor structure\n")
str(samp_fact)
levels(samp_fact)
```
Notice that the factor levels do NOT match the factor labels!

We can also convert samp1 into an ordered factor.
```{r ordered factor numeric}
samp_ord <- factor(samp, ordered = T)
str(samp_ord)
```

## Coercion with as.factor, as.ordered

If we are okay with default settings, **as.factor** can coerce a vector into a factor.
```{r as.factor}
as.factor(grades)
cat("\n") # extra line
as.factor(samp)
```

Similarly, we can use **as.ordered** to coerce a vector into an ordered factor.
```{r as.ordered}
as.ordered(grades)
cat("\n") # extra line
as.ordered(samp)
```

**NOTES**

* It makes sense to designate data as a factor or ordered factor if you have categorical or ordinal data, respectively. But you should always check to be sure there are fewer factor levels than observations, and some R functions have problems with factors that have too many levels.

# Dataframes

Data frames consist of multiple vectors of equal length bound together. Each column in a data frame is like vector in that it can only hold one type of data. But the different columns in a data frame can, and usually do, hold different kinds of data. You will do most of your analysis in R working with data frames.

## data.frame function

We will create a student dataframe to illustrate how the **data.frame** function works.
```{r data.frame 1}
name <- c("Leila", "Vijay", "Shira", "Behrooz")
gender <- factor(c("F", "M", "F", "M"))
age <- c(18, 23, 48, 19)
height <- c(60, 70, 65, 72)
grade <- factor(c("A", "D", "B", "C"), levels = c("F", "D", "C", "B", "A"), ordered = T)

# Now put all of these factors and vectors of same length into a data frame.
students <- data.frame(name, gender, age, height, grade)
str(students)
students

head(students)
```

# column names

## Designate column names with data.frame

Data frame column names are generally variable names. You can designate column names in the **data.frame** function. By default, the names of the vectors used in data.frame become column names, and function names used to create data also become names.

```{r data.frame colnames 1}
# This time set stringsAsFactors=F
students <- data.frame(first.name = name, gender, c(18, 23, 48, 19), height, grade)
students
```

We can designate the third column name as *age* using the following approach.
```{r data.frame colnames 2}
# This time set stringsAsFactors=F
students <- data.frame(first.name = name, gender, age = c(18, 23, 48, 19), height, grade, stringsAsFactors = F)
students
```

## Designate column names with colnames()

We can extract the column names of a data frame using the **colnames** function.
```{r colnames 1}
# extract column names using colnames() function
colnames(students)
```

We can also assign column names of a data frame using the **colnames** function.
```{r colnames() 2}
# assign column names using colnames() function
colnames(students) <- c("First.name", "Gender", "Age", "Height", "Grade")
students
```

## Address data using column names

Column names can be used to address a particular column in a data frame. (All column names in one data frame must be unique.) You can address a column using the format: *dataframe$column.name*.
```{r extract column by name}
students$Age
```

Change data type for one column by name.
```{r manipulate column by name}
students$Age <- as.integer(students$Age)
students
```

Note that we normally wouldn't designate **Age** as an integer. Even though our Age values are all integers, we know that Age is a continuous variable.

# Addressing vectors

Addressing allows for concise data reshaping, extraction and assignment in R. It takes some time to master addressing as there are multiple ways to address and these ways act somewhat differently on different kinds of objects.

## extraction - indexing w/ Brackets []

Square brackets are used to address and extract specific locations within a vector.
```{r vector 1}
vect <- c(1, -2, 4, 6, 2.2, 12) # create vector data object
vect # print data object to the console
cat("\nextract single position (3)\n")
vect[3]

cat("\nextract sequence of positions 2:4\n")
vect[2:4]

cat("\nextract positions 1,3,5.  c(1,3,5)\n")
vect[c(1, 3, 5)]

cat("\nextract everything except position 4  c(-4)\n")
vect[c(-4)]

cat("\nextract everything except positions 2,3,4  -c(2:4)\n")
vect[-c(2:4)]
```

Notice that we extract **values** held at the addressed positions in the vector data object.

Repeat with a character vector.
```{r pets}
pets <- c("hamster", "cat", "dog") # create vector data
pets

cat("\nsingle position (3)\n")
pets[3]

cat("\nsequence of positions 1:2\n")
pets[1:2]

cat("\nselected positions c(1,3)\n")
pets[c(1, 3)]

cat("\nselected positions to remove c(-2)\n")
pets[c(-2)]
```

Notice that element names remain with subsetting.
```{r vector name 2}
# Create a vector with my height and weight in SI units (m, kg)
vect <- c("height (m)" = 1.75, "weight(kg)" = 70)
vect

cat("\nselected positions (2)\n")
vect[2]

vect <- vect[2]
vect
length(vect)
```

## assignment - indexing w/ Brackets []

Addressing locations with brackets can also be used to make assignments, to store data in particular locations within a vector.
```{r vector 1 assignment}
vect <- c(1, -2, 4, 6, 2.2, 12) # create vector data object
vect # print data object to the console
cat("\nassign 0 to single position (3)\n")
vect[3] <- 0
vect

vect <- c(1, -2, 4, 6, 2.2, 12) # create vector data object
cat("\nassign 0.9 to sequence of positions 2:4\n")
vect[2:4] <- .9
vect

vect <- c(1, -2, 4, 6, 2.2, 12) # create vector data object
cat("\nassign -5 to selected positions c(1,3,5)\n")
vect[c(1, 3, 5)] <- -5
vect
```

Repeat assignment with a character vector.
```{r pets assignment}
pets <- c("hamster", "cat", "dog") # create vector data
pets

cat("\n ferret to single position (3)\n")
pets[3] <- "ferret"
pets

pets <- c("hamster", "cat", "dog") # create vector data
cat("\nferret to sequence of positions 1:2\n")
pets[1:2] <- "ferret"
pets

pets <- c("hamster", "cat", "dog") # create vector data
cat("\nferret, monkey to sequence of positions 1:2\n")
pets[1:2] <- c("ferret", "monkey")
pets

pets <- c("hamster", "cat", "dog") # create vector data
cat("\ngerbil, ferret to selected positions 1,3\n")
pets[c(1, 3)] <- c("gerbil", "ferret")
pets

pets <- c("hamster", "cat", "dog") # create vector data
cat("\ngerbil, ferret to all four positions\n")
pets[c(1:4)] <- c("gerbil", "ferret")
pets
```

## Address with logical vectors & []

You can also extract with logical vectors the same length as the vector being addressed inside [].
```{r logical subsetting}
vect <- c(1, -2, 4, 6, 2.2, 12) # create vector data object
vect # print data object to the console
cat("\nc(T,F,T,T,F,T)\n")
vect[c(T, F, T, T, F, T)]

cat("\nc(F,F,T,F,F,F)\n")
vect[c(F, F, T, F, F, F)]

cat("\n")
pets <- c("hamster", "cat", "dog") # create vector data
pets

cat("\nc(T,F,T)\n")
pets[c(T, F, T)]

cat("\nc(F,F,T)\n")
pets[c(F, F, T)]

cat("\nc(F,F,F)\n")
pets[c(F, F, F)]
```

Taking the next step, you can construct a test for vectors values that is then used to extract a subset of data using a logical vector and square brackets.
```{r logical subsetting with test}
vect <- c(1, -2, 4, 6, 2.2, 12) # create vector data object
vect # print data object to the console

test <- vect > 0 # create logical vector
cat("\nvect1 >0\n")
test

cat("\nselect vect1 >0\n")
vect[test]

cat("\nselect vect1 \n")
vect[!test] # ! means 'not' logically, so T -> F, F -> T

test <- vect > 0 # create logical vector
cat("\nvect1 >0\n")
test
cat("\nlength for number of T in logical vector\n")
length(test[test])

cat("\nsum for number of T in logical vector\n")
sum(test)
```

## other addressing issues

What happens with non-standard use of square brackets [].?
```{r}
cat("\n")
pets <- c("hamster", "cat", "dog") # create vector data
pets

cat("\nNotice what we get with []\n")
pets[]

cat("\nNotice what we get with [0]\n")
pets[0]
```

NA always gives NA.
```{r non-standard subsetting}
cat("\n")
pets <- c("hamster", "cat", "dog") # create vector data
pets

cat("\nc(T,NA,F)\n")
pets[c(T, NA, F)]

cat("\nc(NA,2)\n")
pets[c(NA, 2)]
```

# Addressing 2D data objects

Addressing applies to matrices and data frames in much the same way as to vectors.

## Addressing matrices []

As with vectors, matrices can be subsetted, addressed and assigned with square brackets. THe difference is that the square brackets contain two parts **[??,??]**, addressing row and column dimensions.

```{r matrix 4 by 5}
# create a larger matrix using sequenced numbers
mat <- matrix(data = seq(-10, 47, by = 3), nrow = 4, ncol = 5, byrow = T)
dim(mat) <- c(5, 4)
mat

cat("\n3rd row, 2nd column [3,2]\n")
mat[3, 2]

cat("\n1st to 3rd row, 2nd column [1:3,2]\n")
mat[1:3, 2]

cat("\n1st to 3rd row, 2nd and 3rd column [1:3,2:3]\n")
mat[1:3, 2:3]

cat("\n1st 3rd 5th row, 3rd column [c(1,3,5),3]\n")
mat[c(1, 3, 5), 3]
```

As with vectors, we can also remove rows or columns with negative indices. For a given dimension, all indices must be positive or all must be negative.
```{r matrix negative indices}
mat # as above

cat("\nremove 3rd row, remove 2nd column [-3,-2]\n")
mat[-3, -2]

cat("\nremove 1st to 3rd row, 2nd column [-(1:3),2]\n")
mat[-(1:3), 2]

cat("\n1st to 3rd row, remove 2nd and 3rd column [1:3,-(2:3)]\n")
mat[1:3, -(2:3)]

cat("\n1st 3rd 5th row, remove 3rd column [c(1,3,5),-3]\n")
mat[c(1, 3, 5), -3]
```

Notice that with matrices, entire rows or columns are removed with negative indices. We can't remove a single element in a matrix.

As with vectors, leaving an index blank means all rows or all columns are retained.
```{r matrix blank indices}
mat # as above

cat("\nremove 3rd row, keep all columns [-3,]\n")
mat[-3, ]

cat("\nkeep all rows, select 2nd column [,2]\n")
mat[, 2]

cat("\nkeep all rows, remove 2nd and 3rd column [,-(2:3)]\n")
mat[, -(2:3)]
```

We can also *assign* values  to specific locations with addressing brackets [].
```{r matrix assign values}
mat # as above

cat("\nassign 0 to 2nd row, 3rd column\n")
mat3 <- mat
mat3[2, 3] <- 0
mat3

cat("\nassign 99 to 2nd column\n")
mat3 <- mat2
mat3[, 2] <- 99
mat3

cat("\nassign new sequence to 3rd row\n")
mat3 <- mat2
mat3[3, ] <- c(0:-3)
mat3
```

Notice that as with vectors, we can assign a vector of values to a range of locations (e.g. sequence 0, -1, -2, -3 as shown above).  

Change the c(0:-3) sequence end point to -1 **c(0:-1)** and rerun to see what happens. No try for -2 and -4.

## addressing dataframes

Data frames can be subsetted similar to matrices. First recreate earlier data frame.
```{r}
name <- c("Leila", "Vijay", "Shira", "Behrooz")
gender <- factor(c("F", "M", "F", "M"))
age <- c(18, 23, 48, 19)
height <- c(60, 70, 65, 72)
grade <- factor(c("A", "D", "B", "C"), levels = c("F", "D", "C", "B", "A"), ordered = T)

# Now put all of these factors and vectors of same length into a data frame.
students <- data.frame(name, gender, age, height, grade, stringsAsFactors = F)
students
```

Now we can extract a subset with bracket addressing.
```{r data.frame subset 1}
students

cat("\n2nd row, 3rd column (Vijay's age)\n")
students[2, 3]

cat("\n1st row (Leila info)\n")
students[1, ]

cat("\n3rd row w/o grade (Shira)\n")
students[3, -5]
```

And we can assign data to a dataframe position
```{r data.frame subset assign}
students

cat("\n2nd row, 3rd column (change Vijay's age from 23 to 32)\n")
students2 <- students
students2[2, 3] <- 32
students2
```

## subsetting matrices or data frames using logical vectors

We can also extend the way we extracted subsets of vectors with logical vectors to data frames.
```{r}
students <- data.frame(name, gender, age, height, grade, stringsAsFactors = F)
students # original data frame

cat("\n1st and 3rd row [c(T,F,T,F),]\n")
students[c(T, F, T, F), ]

cat("\n1st and 5th column [,c(T,F,F,F,T)]\n")
students[, c(T, F, F, F, T)]

students[students$age < 25, ]
```

Note that the logical vector length has to match the dimension size. We can use the same approach to address matrices.

## Address variables by name

We often address dataframe variables by name using the '$' symbol.
```{r data.frame subset by $}
students

cat("\nstudents$age\n")
students$age

cat("\nstudents$grade\n")
students$grade

# we can combine $ and []
cat("\nnames, 2nd and 3rd $name[2:3]\n")
students$name[2:3]

cat("\nrows 1:3 $age\n")
students[1:3, ]$age
```

# Reordering vectors, matrics and dataframes

Addressing can be used to reorder without necessarily reducing data.

## order vector data

Note that square brackets can also be used to reorder data.
```{r reorder c}
vect <- c(1, -2, 4, 6, 2.2, 12) # create vector data object
vect # print data object to the console

cat("\nc(3,1,2,4,5,6)\n")
vect[c(3, 1, 2, 4, 5, 6)]

cat("\n")
pets <- c("hamster", "cat", "dog") # create vector data
pets

cat("\nc(3,2,1)\n")
pets[c(3, 2, 1)]
```

The **order** function extracts the alphanumeric order of values in a vector.
```{r order}
vect <- c(1, -2, 4, 6, 2.2, 12) # create vector data object
vect # print data object to the console
order(vect)

cat("\n")
pets <- c("hamster", "cat", "dog") # create vector data
pets
order(pets)

cat("\n")
order(pets, decreasing = T)
```

You can use **order** to extract the order, then sort using addressing. There is no base 'sort' function in R, though there are sorting functions available in data wrangling packages.
```{r sort}
vect <- c(1, -2, 4, 6, 2.2, 12) # create vector data object
vect # print data object to the console
cat("\nsort increasing\n")
order(vect)
vect[order(vect)]

cat("\nsort decreasing\n")
order(vect, decreasing = T)
vect[order(vect, decreasing = T)]

cat("\n\n")
pets <- c("hamster", "cat", "dog") # create vector data
pets
cat("\nsort increasing\n")
order(pets)
pets[order(pets)]

cat("\nsort decreasing\n")
order(pets, decreasing = T)
pets[order(pets, decreasing = T)]
```

## order matrices or dataframes

Similarly, we can reorder matrices and dataframes along either (or both) dimensions.
```{r data.frame grades reorder}
students <- data.frame(name, gender, age, height, grade, stringsAsFactors = F)
students # original data frame

cat("\nlast row first [c(4,1,2,3),]\n")
students[c(4, 1, 2, 3), ]

cat("\nput grade in 2nd column [,c(1,5,2,3,4)]\n")
students[, c(1, 5, 2, 3, 4)]
```

And we can sort using the order or one (or more) columns.
```{r data.frame grades sort}
students <- data.frame(name, gender, age, height, grade, stringsAsFactors = F)
students # original data frame

cat("\ngrade order\n")
grade.ord <- order(students$grade, decreasing = T)
grade.ord

# sort by grade order
students[grade.ord, ]

# sort by grade order, same result
students[order(students$grade, decreasing = T), ]

# sort by gender, then age
students[order(students$gender, students$age), ]
```
