---
title: "Data Import and Export and Advanced Subsetting and Addressing"
subtitle: "Data in R, Spring 2026"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


###
knitr::opts_knit$set(root.dir = "C://Users//ottenheimew//Documents//Econ396-Rep//ex directory")
```

#Working Directories



Use getwd and setwd
setwd must contain a file path

```{r}
getwd()
```


```{r}
#setwd("C://Users//ottenheimew//Documents//Econ396-Rep//ex directory")

write.csv2(1:9, "C:\\Users\\ottenheimew\\Documents\\Econ396-Rep\\ex directory\\subsetfolder\\filename3.csv")
```



# Import & Export - Intro

The two R functions commonly used to read external data into data frames are **read.table** and **read.csv**. Each of these has a paired function that writes data frames to an external data file, **write.table** and **write.csv**.

## write.table

We can create a dataframe with student information.
```{r create students dataframe}
name <- c("Matt", "Sid", "Sienna", "Heather")
gender <- factor(c("F","M","F","M"))
age <- c(18,23,48,19)
height <- c(60,70,65,72)
grade <- factor(c("A", "D", "B","C"), levels=c("F","D","C","B","A"), ordered=T)

# Now put these vectors into a data frame.
students <- data.frame(name, gender, age, height, grade,stringsAsFactors=F)
students
row.names(students)
str(students)
```

Now we can export this to an operating system file outside the R environment using **write.table()**.
```{r students write.table}
# the write.table command writes an R data frame out to an external data file.
# in this case we separate with a tab (tab-delimited)
# row names are not exported, we normally don't want these
# quotes are turned off
write.table(students, "students.txt", sep="\t", row.names=F, quote=F)
```


## read.table()

Now we can read this data into R using **read.table**.
```{r read.table}
# By default, read.table recognizes many different separators
#       spaces, tabs, newlines or carriage returns
# header=T tells R that the first row contains column names
students <- read.table("students.txt", header=T)
students
str(students)
```



The variable **gender** was read in as a character, we can coerce to factor with the following code. 
```{r eval = FALSE}
students$gender = as.factor(students$gender)
students
str(students)
```

Notice that grade is NOT a factor. 

```{r fix grade}
# By default, read.table recognizes many different separators
students$grade <-as.factor(students$grade)
str(students)

# Is there a problem? How will you fix it?

```

Now try this. Is there an issue? How will you fix it?

```{r read and write check}
name <- c("Matt Baker", "Sid Chaudhary", "Sienna Yourek", "Heather Wood")
gender <- factor(c("F","M","F","M"))
age <- c(18,23,48,19)
height <- c(60,70,65,72)
grade <- factor(c("A", "D", "B","C"), levels=c("F","D","C","B","A"), ordered=T)
# Now put these vectors into a data frame
students <- data.frame(name, gender, age, height, grade,stringsAsFactors=F)
# Now write the data
write.table(students, "students.txt", sep="\t", row.names=F, quote=F)
# Now read the data. What happens?
students <- read.table("students.txt")
students

```

## read.csv, write.csv

Now we can repeat the excercise above using **read.csv** and **write.csv**.
```{r write.csv}
# the write.csv function writes an R data frame out to an external csv data file.
# row names are not exported
# quotes are turned off
name <- c("Matt", "Sid", "Sienna", "Heather")
students <- data.frame(name, gender, age, height, grade,stringsAsFactors=F)

write.csv(students, "students.csv", row.names=F, quote=F) 
```

Now we can read this data into R using **read.csv**.
```{r read.csv}
# read.csv handles row names and column names automatically
students <- read.csv("students.csv", stringsAsFactors = F)
students
```

Notice that **name** is now a character variable as desired, but **gender** is also a character vector when we want it to be a factor.

In general, it is convenient to work with csv files as they provide efficient data storage and can be recognized by many software packages.

## Datasets on the internet 

It is easy to read in data from the internet. Simply provide the URL address where you would normally put the file name for **read.table** or **read.csv**. We will also introduce the**head** and **summary** commands to help look at a larger data frame.

```{r read.table wine}
wine <- read.table('http://archive.ics.uci.edu/ml/machine-learning-databases/wine/wine.data',sep=",")
head(wine)  # head function looks at the first six rows
```

In this case, the data provided includes no varible names. We can find the variable names and data description at:  archive.ics.uci.edu/ml/machine-learning-databases/wine/wine.names.

We need to add to add the variable names and set the first variable as a factor (wine class = 1, 2, or 3).

```{r modify wine}
colnames(wine) <- c("class","alcohol","malic.acid","ash","alkalinity.ash","mg","tot.phenol","flavanoids","nonflav.phenol","proanthocyanins","col.intensity","hue","OD280.OD315","proline")
wine$class<-as.factor(wine$class)
head(wine)
summary(wine)
```

We can see that the wine data is separated by commas. Can we import with **read.csv**?
```{r read.csv wine}
# note that we don't include the sep parameter for read.csv
wine <- read.csv('http://archive.ics.uci.edu/ml/machine-learning-databases/wine/wine.data')
head(wine)  # head function looks at the first six rows

```

The **read.csv** function expects a header row, so the first row of data are turned into headers.

We can rerun this with header=F to import the data properly. Then we would still need to label the columns and turn the first variable into a factor as before.
```{r read.csv wine 2}
# note that we don't include the sep parameter for read.csv
wine <- read.csv('http://archive.ics.uci.edu/ml/machine-learning-databases/wine/wine.data', header=F)
head(wine)  # head function looks at the first six rows
```

**Notes**

* If data is separated by commas, you can use read.table or read.csv. Otherwise, read.table is the more general function.
* Just about any common data file can be imported in R, either using a base function or a specialized package. THis would include spatial data, image files, excel files, etc...
* R isn't designed for internal data storage. Import data, work on it, then export processed data for external storage. Or increasingly, you can work in R on data that is stored and managed elsewhere (the data is never actually imported into the R environment but through various front end packages data objects can be created that reference external data.)

# Lists in R

Lists are ubiquitous in R, and the most challenging kind of data object to master. Think of a list as a special kind of vector where every element is a data object rather than a value (e.g. vector, factor, matrix, dataframe). You can even have lists within lists!

Many functions generate results in the form of a list. For example, when you run a regression model, the results (r squared, predictor significance, coefficients, etc..) will be combined into a list with a specific format that other functions can recognize.

## create lists

WE can create a list with the **list()** function.
```{r list function 1}
# we can construct a list by stringing together multiple objects
a = c(2, 3, 5) 
b = c("aa", "bb", "cc", "dd", "ee") 
c = c(T, F, T, F, F)
my_list = list(a, b, c, 3)   # contains copies of a, b, c
my_list

cat("\nstructure of my_list\n")
str(my_list)
```

Notice:

* A list may include different data types.
* Each position within a list is noted with double brackets [[]].

We can also name elements in a list.
```{r list function 2}
my_list = list(a=c(2,3,5), b=c("aa","bb","cc","dd","ee"),c=c(T,F,T,F,F), d=3)  
my_list

cat("\nstructure of my_list\n")
str(my_list)

cat("\nnames(my_list)\n")
names(my_list)

cat("\nyou are kidding right\n")
names(my_list) <- c("you","are","kidding","right")
str(my_list)
```

A list item can be a matrix, dataframe, array, etc...
```{r list function 3}
my_list = list(a=c(2,3,5), b=c("aa","bb","cc","dd","ee"), c=c(T,F,T,F,F), d=matrix(seq(1:6),2,3)) 
# readability
my_list = list(a = c(2,3,5), 
               b = c("aa","bb","cc","dd","ee"), 
               c = c(T,F,T,F,F),
               d = matrix(seq(1:6),2,3)) 

my_list

cat("\nstructure of my_list\n")
str(my_list)
```


## List within a list

Another example, with a list within a list.
```{r list within list}
children <- list(names=c("Leila","Elsina"),ages=c(15,12))
children

dave_list <- list(name="dave", job="prof", children=children)
dave_list

cat("\nagain\n")
dave_list_list <- list(name="dave", 
                       job="prof",
                       children=list(names=c("Leila","Elsina"),
                                     ages=c(15,12)))
dave_list_list

str(dave_list_list)
```


```{r list within list nonames}
dave_list <- list("dave", "prof", list(c("Leila","Elsina"),c(13,10)))
dave_list

str(dave_list)
```

## complex lists to organize function outputs

We can generate and examine the output data object from a simple regression.
```{r lm output object}
# generate and examine data
x <- seq(1:30)
y <- 2*seq(1:30)+rnorm(n=30,mean=0,sd=2)
plot(x,y)

# regression with result in xy.lm
xy_lm <- lm(y~x)
xy_lm

cat("\nnow look at str(xy_lm)\n")
str(xy_lm)


```

When a regression model output *lm* object has been created, different R functions know how to handle and use this object.
```{r lm output object 2}
cat("\nDifferent functions know how to handle this complex object, structured in a list\n")

cat("\nclass(xy.lm)------------------------------\n")
class(xy_lm)

cat("\nsummary(xy.lm)----------------------------\n")
summary(xy_lm)

plot(xy_lm)
```

# Lists - addressing

Lists are addressed just like vectors, though it can get more complicated as the items *within* a list can also be addressed.

## Single bracket extraction

### Simple list

We can use single brackets **[]** to obtain subsets of this list.
```{r simple list subsets 1a}
my_list = list(a=c(2,3,5),b=c("aa","bb","cc","dd","ee"),c=c(T,F,T,F,F), d=3)  
my_list

cat("\n> my_list[2]\n\n")
my_list[2]

cat("\n> str(my.list1[2])\n\n")
str(my_list[2])
```

```{r simple list subsets 1b}
my_list[2:4]
```

```{r simple list subsets 1c}
my_list[-3]
```

```{r simple list subsets 1d}
my_list[c(T,F,F,T)]
```

### extracting from more complex list

Repeat **[]** subsetting with more complex list.
```{r simple list subsets 2a}
dave_list <- list(name="dave", job="prof", children=list(names=c("Leila","Elsina"),ages=c(15,12)))
dave_list

cat("> dave_list[2]\n\n")
dave_list[2]
dave_list$job
```

Notice that even when extracting a single value, it is still considered a list. *prof* is the single value in a vector within at list with one element *job*.

More examples
```{r simple list subsets 2b}

cat("> dave_list[3]\n\n")
dave_list[3]

cat("> dave_list[-2]\n\n")
dave_list[-2]

cat("> dave_list[c(T,F,T)]\n\n")
dave_list[c(T,F,T)]
```
All of these results are lists.

## Double bracket extraction

The result of subsetting using a single bracket **[]** is always a list, even if you are extracting a single vector element. If you want to extract the element outside of a list, use double brackets **[[]]**.

```{r list double brackets 1}
my_list  # print for comparison

cat("\n> my_list[2]\n\n")
my_list[2]

cat("> str(my_list[2])\n\n")
str(my_list[2])

cat("\n\n> my_list[[2]]\n\n")
my_list[[2]]

cat("\n> str(my_list[[2]])\n\n")
str(my_list[[2]])
```

Double brackets for first list, 1st item.
```{r list double brackets 2a}
dave_list  # print for comparison

cat("##### Single bracket []\n\n")
cat("> dave.list[1]\n\n")
dave_list[1]
cat("\n> str(dave.list[1])\n\n")
str(dave_list[1])

cat("\n\n##### Double bracket []\n\n")
cat("> dave.list[[1]]\n\n")
dave_list[[1]]
cat("\n> str(dave_list[[1]])\n\n")
str(dave_list[[1]])
```

Double brackets for third list
```{r list double brackets 2b}
dave_list  # print for comparison

cat("##### Single bracket []\n\n")
cat("> dave.list[3]\n\n")
dave_list[3]
cat("> str(dave_list[3])\n\n")
str(dave_list[3])

cat("\n\n##### Double bracket []\n\n")
cat("> dave_list[[3]]\n\n")
dave_list[[3]]
cat("> str(dave.list[[3]])\n\n")
str(dave_list[[3]])
```

## $ addressing similar to [[]]

You can also use the **$** symbol for named lists and get an equivalent result to double brackets.
```{r list dollar sign}
dave_list  # print for comparison

cat("> dave_list$name\n\n")
dave_list$name

cat("\n> dave_list$children\n\n")
dave_list$children
```

## burrowing down into a list

We can address specific data locations within elements of a list, or a list within a list, etc...
```{r list burrow down my.list1}
my_list  # print for comparison

#extract third position from 2nd element of list
cat("\n> my.list1[[2]][3]\n")
my_list[[2]][3]

#This also works
cat("\n> my.list1$b[3]\n")
my_list$b[3]

#Single brackets do NOT work for burrowing down
cat("\n> my.list1[2][3]\n")
my_list[2][3]
```

The last expression looks for the third element in the *list* extracted by my.list1[2], and there is no third element!

Now we can repeat for our second example.
```{r list burrow down dave_list}
dave_list  # print for comparison

cat("\n> my_list[[3]][1]\n")
dave_list[[3]][1]

str(dave_list[[3]][1])

cat("\n> dave_list[[3]][[1]]\n")
dave_list[[3]][[1]]
str(dave_list[[3]][[1]])

cat("\n> dave_list[[3]][[1]][2]\n")
dave_list[[3]][[1]][2]

cat("\n> dave.list$children$names[2]\n")
dave_list$children$names[2]
```

## List addressing to replace values

We need to address a value location to insert a value.
```{r list burrow down dave.list replace 1}
dave_list2 <- dave_list  
dave_list2      # print for comparison

cat("\n> dave_list2[[1]][1] <- 'David'\n")
dave_list2[[1]][1] <- 'David'
dave_list2
```

We can burrow down to replace.
```{r list burrow down dave.list replace 2}
dave_list2 <- dave_list  
dave_list2      # print for comparison

cat("\n> dave_list2[[3]][[1]][2] <- 'Elsie'\n")
dave_list2[[3]][[1]][2] <- 'Elsie'
dave_list2
```

It is easy to make a mistake in addressing a list, so we often prefer to use names.
```{r list burrow down dave.list replace 3}
dave_list2 <- dave_list  
dave_list2      # print for comparison

cat("\n> dave.list2$children$names[1] <- 'Layla'\n")
dave_list2$children$names[1] <- 'Layla'
dave_list2
```


##cbind() and rbind() dataframes

The **cbind()** function operates slightly differently for dataframes.

Let's recreate the student dataframe discussed previously.
```{r student dataframe cbind 1}
name <- c("Leila", "Vijay", "Shira", "Behrooz")
gender <- factor(c("F","M","F","M"))
age <- c(18,23,48,19)
height <- c(60,70,65,72)
grade <- factor(c("A", "D", "B","C"), levels=c("F","D","C","B","A"), ordered=T)

# Now put all of these factors and vectors of same length into a data frame.
students <- data.frame(name, gender, age, height, grade)
students
```

Now let's add a variable for major.
```{r student dataframe cbind2}

cbind(students,c("soil sci", "hort", "econ", "entomology"))

cbind(students,major = c("soil sci", "hort", "econ", "entomology"))

```

Now let's examine the difference between the cbind and data.frame functions.
```{r student dataframe cbind3}

major <- c("soil sci", "hort", "econ", "entomology")

cbind(students,major)

data.frame(students,major)

# cbind applied to dataframes is really just a wrapper function for data.frame. 
# You are better off using data.frame due to greater functionality.
```

It is good practice to use data.frame instead of cbind to join columns of data in data frames.

We can also use rbind() with data frames to add rows (observations).
```{r student dataframe rbind1}
students <- data.frame(name, gender, age, height, grade)
students

students2 <- data.frame(name="Bob", gender="M", age=22, height=69, grade=factor(c("F"), levels=c("F","D","C","B","A"), ordered=T))
students2

rbind(students, students2)
```

Be careful about how R will coerce data if the data types in a given column don't match.
```{r student dataframe rbind2}

students2 <- data.frame(name="Bob", gender="M", age=22, height=69,grade="F")
students2

rbind(students, students2)
str(rbind(students, students2))
```

In this case, R converted the character "F" in *student2* to the ordinal value "F" in the first dataframe.

What happens if column names don't match?
```{r student dataframe rbind3}

# notice Age is capitalized for students2 vs lower case for students.
students2 <- data.frame(name="Bob", gender="M", age=22, height=69, grade="F")
students2

rbind(students, students2)
```

# Advanced addressing - dataframes

read wine
```{r read wine}
wine.dat <- read.csv('http://archive.ics.uci.edu/ml/machine-learning-databases/wine/wine.data',header=F)
colnames(wine.dat) <- c("class","alcohol","malic.acid","ash","alkalinity.ash","mg","tot.phenol","flavanoids","nonflav.phenol","proanthocyanins","col.intensity","hue","OD280.OD315","proline")
wine.dat$class<-as.factor(wine.dat$class)
summary(wine.dat)
str(wine.dat)
```

Now get subset with proline < 1500.

```{r}
wine.dat2 <- wine.dat[wine.dat$proline < 1500, ]
summary(wine.dat2)
```

How do we select class == 1?
```{r eval=FALSE}
wine.dat3 <- wine.dat[wine.dat$class==1,]
summary(wine.dat3$class)

cat("\n")
wine.dat7 <- wine.dat[wine.dat$class=="1",]
summary(wine.dat3$class)

cat("\n")
wine.dat3 <- wine.dat[wine.dat$class==factor("1", levels=c("1","2","3")),]
summary(wine.dat3$class)

cat("\n")
#wine.dat3 <- wine.dat[wine.dat$class==factor("1"),]
#summary(wine.dat3$class)

```

## unique or duplicated vector values

The unique() and duplicated() functions are often useful.
```{r vectors unique duplicated}
vect1 <- c(0,1,1,2,1,3,4)
vect1

#which positions are duplicated?
cat("\nduplicated positions\n")
duplicated(vect1)


#remove duplicated positions
cat("\nremove duplicated positions\n")
vect1[!duplicated(vect1)]

# can also be accomplished with unique()
cat("\nunique values\n")
unique(vect1)
```

These functions can be extended to dataframes.

## unique or duplicated rows

Create students dataframe with duplicate rows.
```{r}
name <- c("Leila", "Vijay", "Shira", "Behrooz")
gender <- factor(c("F","M","F","M"))
age <- c(18,23,48,19)
height <- c(60,70,65,72)
grade <- factor(c("A", "D", "B","C"), levels=c("F","D","C","B","A"), ordered=T)
students <- data.frame(name, gender, age, height, grade)
students2 <- rbind(students,students[2,])
students2
```

Now apply duplicates, unique. Note that these are applied by default to entire rows.
```{r dataframe dup unique}
# duplicated is applied row-wise to dataframes

cat("\nduplicated rows in students2\n")
duplicated(students2)

# remove duplicated rows in students2
students2[!duplicated(students2),]

# select unique rows in students2
unique(students2)
```

Where duplicated() comes in useful is when you are only concerned about duplicates in certain columns of a dataframe, and want to remove rows where those columns are duplicated.
```{r dataframe dup unique 2}
#duplicated is applied row-wise to dataframes
students3 <- students2
students3$grade[2] <- "C"
students3

cat("\nduplicated rows in students3[,1:4]\n")
duplicated(students3[,1:4])

# remove duplicated rows in students3
students3[!duplicated(students3[,1:4]),]

#unique rows in students3
unique(students3)

#unique rows in unique(students3[,1:4])
unique(students3[,1:4])
```

Duplicated removes the row where the first four columns are duplicated, whereas unique can only be applied to an entire row.

##merge()

We can use the merge() function to join two data frames connected by one or more index columns. This will be addressed in greater depth in data wrangling.
```{r student dataframe cbind}
students3

name2 <- c("Leila","Vijay","Bob","Behrooz")
major <- c("soils","hort","plant path","econ")
students.m <- data.frame(name=name2,major)
students.m

#default merge, assume index columns have same names (name column)
mergedstudents<- merge(students3,students.m)

#change merge parameters. all.x keep all x dataframe rows? all.y keep all y dataframe rows? Sort result?
newmergedstud <- merge(students3,students.m, all.x=T, all.y=F, sort=F)
```

#Addressing data in functions

There are multiple ways to use data in a function.

First let's create a dataframe to use as an example.
```{r create dataframe}
mydat <- data.frame(x.var=c(1,5,7,9, 10),y.var=c(2,10,20,90,85))
mydat
```

##explicitly address all variables

plot variables addressed.
```{r data explicit}
lm(mydat$y.var ~ mydat$x.var)
plot(mydat$x.var, mydat$y.var)
```

plot variables addressed with column indexing
```{r}
mydat[,1]
lm(mydat[,2] ~ mydat[,1])
plot(mydat[,1], mydat[,2])
```

Plot with data =
  ```{r data equals}
#lm(y.var ~x.var)
#lm(y.var~x.var, data=mydat)
#plot(x.var, y.var, data=mydat)
```

Doesn't work for all functions.

Use with() function.
```{r with}
with(mydat, lm(y.var ~x.var))
with(mydat, plot(x.var, y.var))
```

Use attach data (warning, be careful, detach after use).
```{r attach}
attach(mydat)
lm(y.var ~x.var)
plot(x.var, y.var)
detach(mydat)
```